---
title: "scToolkit: single cell downstream analyses"
author: "Kevin Blighe, Steven Hargreaves, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('umap')`"
output:
  github_document:
    toc: false
    toc_depth: 3
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{scToolkit: single cell downstream analyses}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

...

```{r, echo=FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

# Installation

## 1. Download the package from Bioconductor

```{r getPackage, eval=FALSE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
    BiocManager::install('scToolkit')

```

Note: to install development version:

```{r getPackageDevel, eval=FALSE}

  devtools::install_github('kevinblighe/scToolkit')

```

## 2. Load the package into R session

```{r Load, message=FALSE}

  #library(scToolkit)

  library(R.utils)
  sourceDirectory('R')

```


# Quick start

...

```{r}

  filelist <- list.files(
    path = "FCS/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)

  metadata <- data.frame(
    file = filelist,
    group = c('PB1', 'HD', 'PB2'),
    row.names = filelist,
    stringsAsFactors = FALSE)

  require(flowCore)
  sct <- processFCS(
    files = filelist,
    metadata = metadata,
    transformation = FALSE,
    downsample = 5)

```

## FACS-like plots for pairwise marker comparison 

...

```{r ex1, fig.height = 8.5, fig.width = 7, fig.cap = "..."}

 require(RColorBrewer)
 require(hexbin)
 require(lattice)

 facsplot(
   data = sct,
   marker1 = 'Ho165Di',
   marker2 = 'Nd144Di',
   log2 = TRUE,
   bins = 200)

```

## Perform principal components analysis

...

```{r ex2, fig.height = 7, fig.width = 7, fig.cap = "..."}

  require(PCAtools)
  p <- pca(t(sct$expression), metadata = sct$metadata)

  biplot(p, lab = FALSE, pointSize = 0.5)

```

## Perform UMAP

...

```{r}

  require(umap)
  sct <- performUMAP(sct)

```

# Advanced features

...

## Create a contour plot of the UMAP layout

...

```{r ex3, fig.height = 8.5, fig.width = 18, fig.cap = "..."}

  ggout <- contourplot(sct)

  require(cowplot)
  plot_grid(ggout, ggout,
    labels = c("Unstimulated", "Stimulated"),
    ncol = 2, align = "h", label_size = 24)

```

## Show marker expression across the layout

...


```{r ex4, fig.height = 12, fig.width = 24, fig.cap = "..."}

  par(mfrow = c(2,4))
  markerExpression(
    data = sct,
    markers = sample(colnames(sct$expression), 8))

```

```{r ex5, fig.height = 7, fig.width = 14, fig.cap = "..."}

  par(mfrow = c(1,2))
  markerExpression(
    data = sct,
    markers = sample(colnames(sct$expression), 2))

```


## Find ideal clusters in the UMAP layout via k-nearest neighbuors

...  

```{r ex6, fig.height = 7, fig.width = 7, fig.cap = "..."}

  sct <- clusKNN(
    sct,
    k.param = 15,
    resolution = 0.001,
    algorithm = 2)

  # HD, unstim
  plot(
    sct$layout,
    main = "",
    xlab = "UMAP 1",
    ylab = "UMAP 2",
    xlim = c(min(sct$layout[,1], na.rm = TRUE) - 1, max(sct$layout[,1], na.rm = TRUE) + 1),
    ylim = c(min(sct$layout[,2], na.rm = TRUE) - 1, max(sct$layout[,2], na.rm = TRUE) + 1),
    col = sct$colour,
    pch = ".")
  text(sct$layout, labels = sct$lab, cex = 2.0, col = "black")


```

## plot marker expression per identified cluster

```{r ex7, fig.height = 14, fig.width = 32, fig.cap = ""}

  require(reshape2)
  require(ggplot2)
  markerExpressionPerCluster(
    data = sct,
    ncol = 3,
    cluster = c(1,2,5,9,11,12))

```

## Differential expression comparison between clusters

```{r}

  res <- diffExpression(
    data = sct,
    cluster1 = 5,
    cluster2 = 8)

  res

```

## Determine enriched markers in each cluster and plot the expression signature

```{r}

  markerEnrichment(sct)

```

```{r ex8, fig.height = 10, fig.width = 16, fig.cap = ""}

  require(corrplot)
  plotSignatures(sct)

```

# Acknowledgments

The development of *scToolkit* has benefited from contributions
and suggestions from:

# Session info

```{r}

sessionInfo()

```

# References
