---
title: "scToolkit: single cell dataviz and downstream analyses"
author: "Kevin Blighe, Jessica Timms, Steven Hargreaves, Shahram Kordasti"
date: "`r Sys.Date()`"
package: "`r packageVersion('SingleCellExperiment')`"
output:
  github_document:
    toc: false
    toc_depth: 3
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{scToolkit: single cell dataviz and downstream analyses}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

In the 'single cell' World, which includes flow cytometry, mass cytometry, single-cell RNA-seq (scRNA-seq), and others, there is a need to improve data visualisation of results and to bring analysis capabilities to researchers even from non-technical backgrounds who have some experience in coding. *scToolkit* [@scToolkit] attempts to fit into this space, while also catering for advanced users. Due to the way that it's designed, *scToolkit* also has a 'plug and play' feel, whereby the base storage unit, which is based on *SingleCellExperiment* [@Lun], immediately lends flexibility and compatibility with studies that go beyond *scToolkit*. Additionally, the graphics are generated via the *ggplot* [@Wickham] engine, which means that users can 'add on' features to these to their pleasing.

...

```{r, echo = FALSE, message = FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

  require(flowCore)
  require(SingleCellExperiment)
  require(PCAtools)
  require(umap)
  require(ggplot2)
  require(ggrepel)
  require(cowplot)
  require(reshape2)
  require(scales)
  require(Seurat)
  require(RColorBrewer)
  require(corrplot)

```

# Installation


## 1. Install from GitHub

```{r getPackageGitHub, eval = FALSE}

  devtools::install_github('kevinblighe/scToolkit')

```

## 2. Load the package into R session

```{r Load, eval = TRUE, echo = FALSE, message = FALSE}

  #library(scToolkit)

  load('save/flow.rdata')

  library(R.utils)
  sourceDirectory('R')

```


# Quick start

Here, we use sample data stored as FCS files.

```{r eval = FALSE, echo = TRUE}

  filelist <- list.files(
    path = "FCS/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)
  filelist

  metadata <- data.frame(
    group = c(rep('Healthy', 7), rep('Disease', 11)),
    treatment = gsub('\\.fcs$', '', gsub('FCS\\/\\/[A-Z0-9]*\\ ', '', filelist)),
    row.names = filelist,
    stringsAsFactors = FALSE)
  metadata

  sce <- processFCS(
    files = filelist,
    metadata = metadata,
    transformation = TRUE,
    downsample = 0.85,
    newColnames = paste0('CD', 1:65))

```

One can also create a new object manually using any type of data, including any data-matrix from scRNA-seq produced elsewhere.

```{r eval = FALSE, echo = TRUE}

  # not run

  mat1 <- jitter(matrix(
    MASS::rnegbin(rexp(4000000, rate=.1), theta = 4.5),
    ncol = 20))
  colnames(mat1) <- paste0('CD', 1:ncol(mat1))
  mat2 <- jitter(matrix(
    MASS::rnegbin(rexp(4000000, rate=.1), theta = 2.5),
    ncol = 20))
  colnames(mat2) <- paste0('CD', 1:ncol(mat2))

  metadata <- data.frame(
    group = c('PB1', 'PB2'),
    row.names = c('mat1', 'mat2'),
    stringsAsFactors = FALSE)

  #...

```

## Perform principal component analysis

```{r ex1, fig.height = 7, fig.width = 8, fig.cap = "biplot"}

  p <- pca(assay(sce, 'scaled'), metadata = metadata(sce))
  biplot(p, lab = NULL, pointSize = 0.5, colby = 'treatment', legendPosition = 'right')

  reducedDim(sce, 'PCA') <- p$rotated[,1:20]

```

For more functionality via *PCAtools*, check the vignette: (PCAtools: everything Principal Component Analysis)[https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html]

## Perform UMAP

UMAP can be performed on the entire dataset, if your computer's memory will permit:

```{r eval = FALSE, echo = TRUE}

  sce <- performUMAP(sce)

```

UMAP can also be stratified based on a column in your metadata, e.g., (treated versus untreated samples); however, to do this, we simply recommend creating separate SingleCellExperiment objects from the very start, i.e., the data input stage, and processing the data separately for each group.

We can also perform UMAP on a select number of PC eigenvectors ('dimensions'). *PCAtools* can be used to infer ideal number of dimensions to use via the elbow method and Horn's parallel analysis.

```{r}

  #elbow <- findElbowPoint(p$variance)
  elbow

  #horn <- parallelPCA(assay(sce, 'scaled'))
  horn$n

```

```{r eval = FALSE, echo = TRUE}

  sce <- performUMAP(sce, reducedDim = 'PCA', dims = c(1:horn$n))

``

```{r eval = FALSE, echo = FALSE}

  #save.image('save/flow.rdata')

```


## Create a contour plot of the UMAP layout

```{r ex2, fig.height = 8, fig.width = 16, fig.cap = "contourplot"}

  ggout1 <- contourplot(sce, reducedDim = 'UMAP', subtitle = 'UMAP performed on expression values')
  ggout2 <- contourplot(sce, reducedDim = 'UMAP_PCA', subtitle = 'UMAP performed on PC eigenvectors')

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 24)

```

## Show marker expression across the layout

...


```{r ex3, fig.height = 12, fig.width = 20, fig.cap = "markerExpression"}

  markers <- sample(rownames(sce), 6)
  markers

  ggout1 <- markerExpression(sce,
    markers = markers,
    subtitle = 'UMAP performed on expression values',
    nrow = 1, ncol = 6,
    legendKeyHeight = 1.0)

  ggout2 <-  markerExpression(sce,
    markers = markers,
    reducedDim = 'UMAP_PCA',
    subtitle = 'UMAP performed on PC eigenvectors',
    nrow = 1, ncol = 6,
    legendKeyHeight = 1.0)

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    nrow = 2, align = "l", label_size = 24)

```

## Shade cells by metadata

```{r}

  head(metadata(sce))

  levels(metadata(sce)$group)

  levels(metadata(sce)$treatment)

```

```{r ex4, fig.height = 12, fig.width = 14, fig.cap = "metadataplot"}

  ggout1 <- metadataplot(sce,
    colby = 'group',
    colkey = c(Healthy = 'royalblue', Disease = 'red2'),
    title = 'Disease status',
    subtitle = 'UMAP performed on expression values')

  ggout2 <- metadataplot(sce,
    reducedDim = 'UMAP_PCA',
    colby = 'group',
    colkey = c(Healthy = 'royalblue', Disease = 'red2'),
    title = 'Disease status',
    subtitle = 'UMAP performed on PC eigenvectors')

  ggout3 <- metadataplot(sce,
    colby = 'treatment',
    title = 'Treatment type',
    subtitle = 'UMAP performed on expression values')

  ggout4 <- metadataplot(sce,
    reducedDim = 'UMAP_PCA',
    colby = 'treatment',
    title = 'Treatment type',
    subtitle = 'UMAP performed on PC eigenvectors')

  plot_grid(ggout1, ggout3, ggout2, ggout4,
    labels = c('A','B','C','D'),
    nrow = 2, ncol = 2, align = "l", label_size = 24)

```

## Find ideal clusters in the UMAP layout via k-nearest neighbuors


```{r ex5, fig.height = 8, fig.width = 16, fig.cap = "clusKNN"}

  sce <- clusKNN(sce,
    k.param = 20,
    prune.SNN = 1/15,
    resolution = 0.01,
    algorithm = 2)

  sce <- clusKNN(sce,
    reducedDim = 'UMAP_PCA',
    clusterAssignName = 'Cluster_PCA',
    k.param = 20,
    prune.SNN = 1/15,
    resolution = 0.01,
    algorithm = 2)

  ggout1 <- plotClusters(sce,
    clusterColname = 'Cluster',
    subtitle = 'UMAP performed on expression values',
    caption = paste0('Note: clusters / communities identified via',
      '\nLouvain algorithm with multilevel refinement'))

  ggout2 <- plotClusters(sce,
    clusterColname = 'Cluster_PCA',
    reducedDim = 'UMAP_PCA',
    subtitle = 'UMAP performed on PC eigenvectors',
    caption = paste0('Note: clusters / communities identified via',
      '\nLouvain algorithm with multilevel refinement'))

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 24)

```

## plot marker expression per identified cluster

```{r ex6a, fig.height = 9, fig.width = 16, fig.cap = "markerExpressionPerCluster 1"}

  markerExpressionPerCluster(sce,
    caption = 'Cluster assignments based on UMAP performed on expression values')

```

```{r ex6b, fig.height = 9, fig.width = 18, fig.cap = "markerExpressionPerCluster 2"}

  markerExpressionPerCluster(sce,
    clusters = unique(metadata(sce)[['Cluster_PCA']]),
    clusterVector = metadata(sce)[['Cluster_PCA']],
    markers = sample(rownames(sce), 15),
    axisLabSize = 14,
    nrow = 2, ncol = 4,
    caption = 'Cluster assignments based on UMAP performed on PC eigenvectors')

```

## Differential expression comparison between clusters

```{r}

  c1 <- rownames(subset(metadata(sce), group == 'Disease'))
  c2 <- rownames(subset(metadata(sce), group == 'Healthy'))

  #res <- diffExpression(sce,
  #  cells1 = c1,
  #  cells2 = c2)

  #res

```

## Determine enriched markers in each cluster and plot the expression signature

`r knitr::kable(markerEnrichment(sce, metacluster = 'group'), auto = TRUE)`

`r knitr::kable(markerEnrichment(sce, metacluster = 'treatment'), auto = TRUE)`   


```{r ex7, fig.height = 10, fig.width = 16, fig.cap = "plotSignatures"}

  plotSignatures(sce)

```

# Advanced features

...



# Acknowledgments


# Session info

```{r}

sessionInfo()

```

# References

﻿@scToolkit

@Lun

@Wickham

