---
title: "scToolkit: single cell downstream analyses"
author: "Kevin Blighe, Steven Hargreaves, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('cytotools')`"
output:
  github_document:
    toc: false
    toc_depth: 3
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{scToolkit: single cell downstream analyses}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

...

```{r, echo=FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

# Installation

## 1. Download the package from Bioconductor

```{r getPackage, eval=FALSE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
    BiocManager::install('scToolkit')

```

Note: to install development version:

```{r getPackageDevel, eval=FALSE}

  devtools::install_github('kevinblighe/scToolkit')

```

## 2. Load the package into R session

```{r Load, message=FALSE}

  library(scToolkit)

```


# Quick start

...

```{r}

  filelist <- list.files(
    path = "FCS/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)

  cte <- processFCS(
    files = filelist,
    markersRetain = c("Y89Di","Rh103Di","Sn120Di",
      "I127Di","Xe131Di","Cs133Di","Ba138Di",
      "Ce140Di","Pr141Di","Nd142Di","Nd143Di",
      "Nd144Di","Nd145Di","Nd146Di","Sm147Di,",
      "Nd148Di","Sm149Di"))

```

## FACS-like plots for pairwise marker comparison 

...

```{r ex1, fig.height = 8.5, fig.width = 7, fig.cap = "..."}

 require(RColorBrewer)
 require(hexbin)
 require(lattice)

 facsplot(
   data = cte,
   marker1 = 'Nd142Di',
   marker2 = 'I127Di',
   x1=0, x2=5, y1=0, y2=5,
   cex = 1.0)

```

## Perform principal components analysis

...

```{r ex2, fig.height = 8.5, fig.width = 7, fig.cap = "..."}

  require(PCAtools)
  p <- pca(t(cte),)

  biplot(p, lab = FALSE, pointSize = 1)

```

## Perform UMAP

...

```{r}

  u <- performUMAP(cte)

```

# Advanced features

...

## Create a contour plot of the UMAP layout

...

```{r ex3, fig.height = 8.5, fig.width = 14, fig.cap = "..."}

  ggout <- contourplot(u)

  require(cowplot)
  plot_grid(ggout, ggout,
    labels = c("Unstimulated", "Stimulated"),
    ncol = 2, align = "h", label_size = 24)

```

## Show marker expression across the layout

...


```{r ex4, fig.height = 16, fig.width = 16, fig.cap = "..."}

  par(mfrow = c(4,4))
  markerExpression(u)

```

## Find ideal clusters in the UMAP layout via k-nearest neighbuors

...  

```{r ex5, fig.height = 8.5, fig.width = 7, fig.cap = "..."}

  u <- clusKNN(u)

  # HD, unstim
  plot(
    u$layout,
    main = "HD, unstimulated",
    xlab = "UMAP 1",
    ylab = "UMAP 2",
    #xlim = c(-15, 10),
    #ylim = c(-13, 14),
    col = u$colour,
    pch = ".")
  text(u$layout, labels = u$lab, cex = 2.0, col = "black")


```

## plot marker expression per identified cluster

```{r ex6, fig.height = 12, fig.width = 16, fig.cap = ""}

  markerExpressionPerCluster(u)

```

## Differential expression comparison between clusters

```{r}

  diffExpression(
    data = u,
    cluster1 = 5,
    cluster2 = 8)

```

## Determine enriched markers in each cluster and plot the expression signature

```

  markerEnrichment()

```

```{r ex7, fig.height = 12, fig.width = 16, fig.cap = ""}

  plotSignatures(u)
  clusSignificancePlot(u)

```

# Acknowledgments

The development of *scToolkit* has benefited from contributions
and suggestions from:

# Session info

```{r}

sessionInfo()

```

# References