---
title: "scDataviz: single cell dataviz and downstream analyses"
author: "Kevin Blighe"
date: "`r Sys.Date()`"
package: "`r packageVersion('scDataviz')`"
output:
  github_document:
    toc: false
    toc_depth: 3
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{scDataviz: single cell dataviz and downstream analyses}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

In the single cell World, which includes flow cytometry, mass cytometry, single-cell RNA-seq (scRNA-seq), and others, there is a need to improve data visualisation and to bring analysis capabilities to researchers even from non-technical backgrounds. *scDataviz* [@scDataviz] attempts to fit into this space, while also catering for advanced users. Additonally, due to the way that *scDataviz* is designed, which is based on *SingleCellExperiment* [@Lun], it has a 'plug and play' feel, and immediately lends itself as flexibile and compatibile with studies that go beyond *scDataviz*. Finally, the graphics in *scDataviz* are generated via the *ggplot* [@Wickham] engine, which means that users can 'add on' features to these with ease.

...

```{r, echo = FALSE, message = FALSE}

  library(knitr)
  library(kableExtra)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

# Installation


## 1. Install from GitHub

```{r getPackageGitHub, eval = FALSE}

  devtools::install_github('kevinblighe/scDataviz')

```

## 2. Load the package into R session

```{r Load}

  library(scDataviz)

```


# Quick start

Here, we use sample data stored as FCS files.

```{r readFCS}

  filelist <- list.files(
    path = "FCS/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)
  filelist

  metadata <- data.frame(
    group = c(rep('Healthy', 7), rep('Disease', 11)),
    treatment = gsub('\\.fcs$', '', gsub('FCS\\/\\/[A-Z0-9]*\\ ', '', filelist)),
    row.names = filelist,
    stringsAsFactors = FALSE)
  metadata

  sce <- processFCS(
    files = filelist,
    metadata = metadata,
    transformation = TRUE,
    downsample = 0.85,
    newColnames = paste0('CD', 1:65))

```

One can also create a new object *SingleCellExperiment* object manually using any type of data, including any data-matrix from scRNA-seq produced elsewhere. Import functions for data deriving from popular programs will come, e.g., *Seurat*.


## Perform principal component analysis (PCA)

```{r ex1, fig.height = 7, fig.width = 8, fig.cap = "Perform principal component analysis"}

  p <- pca(assay(sce, 'scaled'), metadata = metadata(sce))
  biplot(p, lab = NULL, pointSize = 0.5, colby = 'treatment', legendPosition = 'right')

```

We can add the rotated component loadings as a new reduced dimensional component to our dataset. Let's just add the first 20 PCs.

```{r addPCAdim}

  reducedDim(sce, 'PCA') <- p$rotated[,1:20]

```

For more functionality via *PCAtools*, check the vignette: [PCAtools: everything Principal Component Analysis](https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html)

## Perform UMAP

UMAP can be performed on the entire dataset, if your computer's memory will permit. Currently it is fixed to work with the 'scaled' assay component of the *SingleCellExperiment* object.

```{r performUMAP}

  sce <- performUMAP(sce)

```

UMAP can also be stratified based on a column in your metadata, e.g., (treated versus untreated samples); however, to do this, I recommend creating separate *SingleCellExperiment* objects from the very start, i.e., from the the data input stage, and processing the data separately for each group.

We can also perform UMAP on a select number of PC eigenvectors. *PCAtools* can be used to infer ideal number of dimensions to use via the elbow method and Horn's parallel analysis.

```{r elbowHorn}

  #elbow <- findElbowPoint(p$variance)
  elbow

  #horn <- parallelPCA(assay(sce, 'scaled'))
  horn$n

```

```{r performUMAP_PCA}

  sce <- performUMAP(sce, reducedDim = 'PCA', dims = c(1:horn$n))

```

At this point, we have done a lot - let's save the dataset.

```{r save, warnings = FALSE, message = FALSE}

  dir.create('save'
  saveRDS(sce, 'save/VignetteData.RDS', compress = TRUE)

```

## Create a contour plot of the UMAP layout

```{r ex2, fig.height = 8, fig.width = 16, fig.cap = "Create a contour plot of the UMAP layout"}

  ggout1 <- contourplot(sce, reducedDim = 'UMAP', subtitle = 'UMAP performed on expression values')
  ggout2 <- contourplot(sce, reducedDim = 'UMAP_PCA', subtitle = 'UMAP performed on PC eigenvectors')

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 24)

```

## Show marker expression across the layout

```{r ex3, fig.height = 12, fig.width = 20, fig.cap = "Show marker expression across the layout"}

  markers <- sample(rownames(sce), 6)
  markers

  ggout1 <- markerExpression(sce,
    markers = markers,
    subtitle = 'UMAP performed on expression values',
    nrow = 1, ncol = 6,
    legendKeyHeight = 1.0)

  ggout2 <-  markerExpression(sce,
    markers = markers,
    reducedDim = 'UMAP_PCA',
    subtitle = 'UMAP performed on PC eigenvectors',
    nrow = 1, ncol = 6,
    legendKeyHeight = 1.0)

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    nrow = 2, align = "l", label_size = 24)

```

## Shade cells by metadata

First, let's take a look inside the metadata that we have.

```{r metadataplot}

  head(metadata(sce))

  levels(metadata(sce)$group)

  levels(metadata(sce)$treatment)

```

```{r ex4, fig.height = 12, fig.width = 14, fig.cap = "Shade cells by metadata"}

  ggout1 <- metadataplot(sce,
    colby = 'group',
    colkey = c(Healthy = 'royalblue', Disease = 'red2'),
    title = 'Disease status',
    subtitle = 'UMAP performed on expression values')

  ggout2 <- metadataplot(sce,
    reducedDim = 'UMAP_PCA',
    colby = 'group',
    colkey = c(Healthy = 'royalblue', Disease = 'red2'),
    title = 'Disease status',
    subtitle = 'UMAP performed on PC eigenvectors')

  ggout3 <- metadataplot(sce,
    colby = 'treatment',
    title = 'Treatment type',
    subtitle = 'UMAP performed on expression values')

  ggout4 <- metadataplot(sce,
    reducedDim = 'UMAP_PCA',
    colby = 'treatment',
    title = 'Treatment type',
    subtitle = 'UMAP performed on PC eigenvectors')

  plot_grid(ggout1, ggout3, ggout2, ggout4,
    labels = c('A','B','C','D'),
    nrow = 2, ncol = 2, align = "l", label_size = 24)

```

## Find ideal clusters in the UMAP layout via k-nearest neighbours

```{r ex5, message = FALSE, fig.height = 8, fig.width = 16, fig.cap = "Find ideal clusters in the UMAP layout via k-nearest neighbours"}

  sce <- clusKNN(sce,
    k.param = 20,
    prune.SNN = 1/15,
    resolution = 0.01,
    algorithm = 2)

  sce <- clusKNN(sce,
    reducedDim = 'UMAP_PCA',
    clusterAssignName = 'Cluster_PCA',
    k.param = 20,
    prune.SNN = 1/15,
    resolution = 0.01,
    algorithm = 2)

  ggout1 <- plotClusters(sce,
    clusterColname = 'Cluster',
    subtitle = 'UMAP performed on expression values',
    caption = paste0('Note: clusters / communities identified via',
      '\nLouvain algorithm with multilevel refinement'))

  ggout2 <- plotClusters(sce,
    clusterColname = 'Cluster_PCA',
    reducedDim = 'UMAP_PCA',
    subtitle = 'UMAP performed on PC eigenvectors',
    caption = paste0('Note: clusters / communities identified via',
      '\nLouvain algorithm with multilevel refinement'))

  plot_grid(ggout1, ggout2,
    labels = c('A','B'),
    ncol = 2, align = "l", label_size = 24)

```

## Plot marker expression per identified cluster

```{r ex6a, fig.height = 8, fig.width = 16, fig.cap = "Plot marker expression per identified cluster1"}

  markerExpressionPerCluster(sce,
    caption = 'Cluster assignments based on UMAP performed on expression values')

```

```{r ex6b, fig.height = 8, fig.width = 18, fig.cap = "Plot marker expression per identified cluster2"}

  markerExpressionPerCluster(sce,
    clusters = unique(metadata(sce)[['Cluster_PCA']]),
    clusterVector = metadata(sce)[['Cluster_PCA']],
    markers = sample(rownames(sce), 15),
    axisLabSize = 14,
    nrow = 2, ncol = 4,
    caption = 'Cluster assignments based on UMAP performed on PC eigenvectors')

```


## Determine enriched markers in each cluster and plot the expression signature

`r knitr::kable(markerEnrichment(sce, metacluster = 'group'), auto = TRUE) %>% kable_styling(full_width = F)

`r knitr::kable(markerEnrichment(sce, metacluster = 'treatment'), auto = TRUE) %>% kable_styling(full_width = F)`


The expression signature is a quick way to visualise which markers are more or less expressed in each identified cluster of cells.

```{r ex7, fig.height = 10, fig.width = 16, fig.cap = "Determine enriched markers in each cluster and plot the expression signature"}

  plotSignatures(sce)

```


# Acknowledgments

* Jessica Timms

* James Opzoomer

* Steven Hargreaves

* Shahram Kordasti

# Session info

```{r}

sessionInfo()

```

# References

﻿@scDataviz

@Lun

@Wickham

