---
title: "scToolkit: single cell downstream analyses"
author: "Kevin Blighe, Steven Hargreaves, Jessica Timms"
date: "`r Sys.Date()`"
package: "`r packageVersion('umap')`"
output:
  github_document:
    toc: false
    toc_depth: 3
fig_width: 7
bibliography: library.bib
vignette: >
    %\VignetteIndexEntry{scToolkit: single cell downstream analyses}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\usepackage[utf8]{inputenc}
---

# Introduction

...

```{r, echo=FALSE}

  library(knitr)
  opts_chunk$set(tidy = FALSE, message = FALSE, warning = FALSE)

```

# Installation

## 1. Download the package from Bioconductor

```{r getPackage, eval=FALSE}

  if (!requireNamespace('BiocManager', quietly = TRUE))
    install.packages('BiocManager')
    BiocManager::install('scToolkit')

```

Note: to install development version:

```{r getPackageDevel, eval=FALSE}

  devtools::install_github('kevinblighe/scToolkit')

```

## 2. Load the package into R session

```{r Load, message=FALSE}

  #library(scToolkit)

  library(R.utils)
  sourceDirectory('R')

```


# Quick start

...

```{r}

  filelist <- list.files(
    path = "FCS/",
    pattern = "*.fcs|*.FCS",
    full.names = TRUE)

  metadata <- data.frame(
    group = c('PB1', 'HD', 'PB2'),
    row.names = filelist,
    stringsAsFactors = FALSE)

  require(flowCore)
  require(SingleCellExperiment)
  sce <- processFCS(
    files = filelist,
    metadata = metadata,
    transformation = TRUE,
    downsample = 0.80,
    newColnames = paste0('CD', 1:52))

```

```{r}

  mat1 <- matrix(
    MASS::rnegbin(rexp(2000000, rate=.1), theta = 4.5),
    ncol = 20)
  colnames(mat1) <- paste0('CD', 1:ncol(mat1))
  mat2 <- matrix(
    MASS::rnegbin(rexp(2000000, rate=.1), theta = 2.5),
    ncol = 20)
  colnames(mat2) <- paste0('CD', 1:ncol(mat2))

  metadata <- data.frame(
    group = c('PB1', 'PB2'),
    row.names = c('mat1', 'mat2'),
    stringsAsFactors = FALSE)

  require(SingleCellExperiment)
  sce <- processData(
    mat = list('mat1' = mat1,
      'mat2' = mat2),
    metadata = metadata,
    transformation = TRUE,
    downsample = 0.80)

```

## FACS-like plots for pairwise marker comparison 

...

```{r ex1, fig.height = 8.5, fig.width = 7, fig.cap = "..."}

 require(RColorBrewer)
 require(hexbin)
 require(lattice)

 facsplot(
   sce = sce,
   marker1 = 'CD3',
   marker2 = 'CD18',
   log2 = TRUE,
   bins = 200)

```

## Perform principal components analysis

...

```{r ex2, fig.height = 7, fig.width = 7, fig.cap = "..."}

  require(PCAtools)
  p <- pca(t(assay(sce, 'scaled')), metadata = rowData(sce))

  biplot(p, lab = FALSE, pointSize = 0.5, colby = 'group')

```

## Perform UMAP

...

```{r}

  require(umap)
  sce <- performUMAP(sce)

```

# Advanced features

...

## Create a contour plot of the UMAP layout

...

```{r ex3, fig.height = 8.5, fig.width = 18, fig.cap = "..."}

  require(ggplot2)
  require(ggrepel)
  ggout <- contourplot(sce)

  require(cowplot)
  plot_grid(ggout, ggout,
    labels = c("Unstimulated", "Stimulated"),
    ncol = 2, align = "h", label_size = 24)

```

## Show marker expression across the layout

...


```{r ex5, fig.height = 7, fig.width = 14, fig.cap = "..."}

  require(reshape2)
  require(scales)
  markerExpression(sce,
    markers = sample(rownames(sce), 8),
    #col = c('green','red'),
    nrow = 2, ncol = 4)

  require(RColorBrewer)
  require(SDMTools)

  par(mfrow = c(1,2))
  markerExpression(sce)

```


## Find ideal clusters in the UMAP layout via k-nearest neighbuors

...  

```{r ex6, fig.height = 7, fig.width = 7, fig.cap = "..."}

  sct <- clusKNN(
    sct,
    k.param = 15,
    resolution = 0.001,
    algorithm = 2)

  # HD, unstim
  plot(
    sct$layout,
    main = "",
    xlab = "UMAP 1",
    ylab = "UMAP 2",
    xlim = c(min(sct$layout[,1], na.rm = TRUE) - 1, max(sct$layout[,1], na.rm = TRUE) + 1),
    ylim = c(min(sct$layout[,2], na.rm = TRUE) - 1, max(sct$layout[,2], na.rm = TRUE) + 1),
    col = sct$colour,
    pch = ".")
  text(sct$layout, labels = sct$lab, cex = 1.0, col = "black")


```

## plot marker expression per identified cluster

```{r ex7, fig.height = 14, fig.width = 32, fig.cap = ""}

  require(reshape2)
  require(ggplot2)
  markerExpressionPerCluster(
    data = sct,
    ncol = 3,
    cluster = c(1,2,5,9,11,12))

```

## Differential expression comparison between clusters

```{r}

  res <- diffExpression(
    data = sct,
    cluster1 = 5,
    cluster2 = 8)

  res

```

## Determine enriched markers in each cluster and plot the expression signature

```{r}

  markerEnrichment(sct)

```

```{r ex8, fig.height = 10, fig.width = 16, fig.cap = ""}

  require(corrplot)
  plotSignatures(sct)

```

# Acknowledgments

The development of *scToolkit* has benefited from contributions
and suggestions from:

# Session info

```{r}

sessionInfo()

```

# References
